---
title: "Plasmid Copy Number Analysis — dPCR"
author: ""
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    theme: cosmo
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(lme4)
library(lmerTest)  # p-values for fixed effects
library(ggplot2)
library(dplyr)
library(tidyr)
library(emmeans)

theme_set(theme_minimal(base_size = 13))
```

## Data Structure

Expected input: a CSV in semi-wide format with columns:

- `strain` (factor, 3 levels)
- `antibiotic` (factor: "none" / "treated" or similar)
- `timepoint` (numeric, e.g. hours)
- `bio_rep` (1 or 2)
- `tech_rep` (1, 2, or 3)
- `plasmid_per_ter` (HEX/FAM ratio from plasmid + terminus duplex)
- `plasmid_per_ori` (HEX/FAM ratio from plasmid + origin duplex)
- `ori_ter` (origin/terminus ratio from the direct comparison plate; 
   same value repeated across tech reps for a given bio sample)

```{r}
#| label: load-data

# ---- Replace this with your actual data file ----
# df <- read.csv("your_data.csv")

# ---- Simulated example data for testing the pipeline ----
set.seed(42)

strains <- c("A", "B", "C")
antibiotics <- c("none", "abx")
timepoints <- c(0, 2, 4, 6)

df <- expand.grid(
  strain = strains,
  antibiotic = antibiotics,
  timepoint = timepoints,
  bio_rep = 1:2,
  tech_rep = 1:3
) |>
  mutate(
    # simulate some biology: copy number varies by strain, responds to abx over time
    true_pcn = 5 +
      (strain == "B") * 3 +
      (strain == "C") * -1 +
      (antibiotic == "abx") * timepoint * 0.5 +
      (strain == "B" & antibiotic == "abx") * timepoint * 0.3,
    # ori/ter increases with growth, varies by condition (sample-level, no tech rep noise)
    ori_ter = 1.5 +
      timepoint * 0.05 +
      (antibiotic == "abx") * -0.1 +
      rnorm(n(), 0, 0.05),
    # bio rep random effect
    bio_effect = rep(rnorm(n() / 3, 0, 0.5), each = 3),
    # observed values with tech rep noise
    plasmid_per_ter = true_pcn + bio_effect + rnorm(n(), 0, 0.3),
    plasmid_per_ori = (true_pcn / ori_ter) + bio_effect / ori_ter + rnorm(n(), 0, 0.25),
  ) |>
  select(-true_pcn, -bio_effect)

# Ensure factors
df <- df |>
  mutate(
    strain = factor(strain),
    antibiotic = factor(antibiotic, levels = c("none", "abx")),
    bio_rep = factor(bio_rep),
    tech_rep = factor(tech_rep),
    # unique biological sample ID for random effect
    sample_id = interaction(strain, antibiotic, bio_rep, drop = TRUE)
  )

glimpse(df)
```

## Sanity Checks

### Ori/Ter consistency across dilutions

If you averaged your two dilutions before building this table, just confirm the
ori/ter values look reasonable. They should be ≥1 (origin is at least as abundant
as terminus) and higher values indicate faster growth.

```{r}
#| label: ori-ter-check

df |>
  distinct(strain, antibiotic, timepoint, bio_rep, ori_ter) |>
  ggplot(aes(x = timepoint, y = ori_ter, colour = strain, shape = antibiotic)) +
  geom_point(size = 3, position = position_dodge(width = 0.3)) +
  labs(y = "Origin / Terminus ratio", x = "Time") +
  ggtitle("Ori/Ter ratio across conditions")
```

### Raw plasmid copy number

```{r}
#| label: raw-data-plot
#| fig-height: 5
#| fig-width: 9

df |>
  ggplot(aes(x = timepoint, y = plasmid_per_ter,
             colour = strain, shape = antibiotic)) +
  geom_point(alpha = 0.4, position = position_jitterdodge(jitter.width = 0.1,
                                                           dodge.width = 0.4)) +
  stat_summary(aes(group = interaction(strain, antibiotic)),
               fun = mean, geom = "line",
               position = position_dodge(width = 0.4)) +
  labs(y = "Plasmid copies / terminus (genome)", x = "Time") +
  ggtitle("Plasmid copy number (terminus-normalised)")
```

## Primary Model: Plasmid per Terminus

Continuous timepoint with three-way interaction. Ori/ter as covariate to account
for growth-rate-associated variation. Random intercept per biological sample.

```{r}
#| label: model-ter

mod_ter <- lmer(
  plasmid_per_ter ~ strain * antibiotic * timepoint + ori_ter +
    (1 | sample_id),
  data = df
)

summary(mod_ter)
```

### Check whether three-way interaction is needed

```{r}
#| label: model-comparison

mod_ter_2way <- lmer(
  plasmid_per_ter ~ (strain + antibiotic + timepoint)^2 + ori_ter +
    (1 | sample_id),
  data = df
)

anova(mod_ter_2way, mod_ter)
```

If the three-way interaction is not significant, drop it and use the simpler
two-way model for cleaner interpretation.

### Diagnostics

```{r}
#| label: diagnostics-ter
#| fig-height: 4
#| fig-width: 8
#| layout-ncol: 2

plot(mod_ter, main = "Residuals vs Fitted")
qqnorm(resid(mod_ter)); qqline(resid(mod_ter))
```

### Estimated marginal means

```{r}
#| label: emmeans-ter
#| fig-height: 5
#| fig-width: 9

# Predictions at each timepoint × strain × antibiotic
# ori_ter held at its mean
em_ter <- emmeans(mod_ter, ~ strain * antibiotic | timepoint,
                  at = list(timepoint = unique(df$timepoint)),
                  lmer.df = "satterthwaite")

em_df <- as.data.frame(em_ter)

ggplot(em_df, aes(x = timepoint, y = emmean,
                  colour = strain, linetype = antibiotic)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = strain),
              alpha = 0.15, colour = NA) +
  labs(y = "Estimated plasmid copies / genome",
       x = "Time") +
  ggtitle("Model estimates: plasmid copy number (terminus-normalised)")
```

### Slopes: rate of change over time

```{r}
#| label: slopes-ter

emtrends(mod_ter, pairwise ~ strain * antibiotic, var = "timepoint",
         lmer.df = "satterthwaite")
```

## Secondary Model: Plasmid per Origin

Same structure, different response. Useful as a consistency check — if plasmid
copy number is genuinely changing, both normalisations should show it, scaled
by the ori/ter ratio.

```{r}
#| label: model-ori

mod_ori <- lmer(
  plasmid_per_ori ~ strain * antibiotic * timepoint + ori_ter +
    (1 | sample_id),
  data = df
)

summary(mod_ori)
```

## Comparing the Two Normalisations

If the two models give qualitatively different answers for a condition, the
discrepancy is informative — it suggests the effect is mediated by (or
confounded with) replication state rather than true copy number regulation.

```{r}
#| label: comparison-plot
#| fig-height: 5
#| fig-width: 10

em_ori <- emmeans(mod_ori, ~ strain * antibiotic | timepoint,
                  at = list(timepoint = unique(df$timepoint)),
                  lmer.df = "satterthwaite") |>
  as.data.frame() |>
  mutate(normalisation = "per origin")

em_both <- em_df |>
  mutate(normalisation = "per terminus") |>
  bind_rows(em_ori)

ggplot(em_both, aes(x = timepoint, y = emmean,
                    colour = strain, linetype = antibiotic)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = strain),
              alpha = 0.12, colour = NA) +
  facet_wrap(~ normalisation, scales = "free_y") +
  labs(y = "Estimated plasmid copies", x = "Time") +
  ggtitle("Plasmid copy number: terminus vs origin normalisation")
```

## Effect of Ori/Ter Covariate

Check whether the ori/ter ratio meaningfully adjusts the estimates. If its
coefficient is near zero or non-significant, growth state isn't confounding
your copy number measurements at these timepoints.

```{r}
#| label: ori-ter-effect

# Coefficient from the terminus model
fixef(mod_ter)["ori_ter"]
summary(mod_ter)$coefficients["ori_ter", ]

# Model without the covariate for comparison
mod_ter_no_ot <- lmer(
  plasmid_per_ter ~ strain * antibiotic * timepoint +
    (1 | sample_id),
  data = df
)

anova(mod_ter_no_ot, mod_ter)
```

## Quick Descriptive Summary

Plain means ± SD by condition and timepoint, averaged over tech reps first.

```{r}
#| label: descriptive

df |>
  group_by(strain, antibiotic, timepoint, bio_rep) |>
  summarise(
    pcn_ter = mean(plasmid_per_ter),
    pcn_ori = mean(plasmid_per_ori),
    ot = mean(ori_ter),
    .groups = "drop"
  ) |>
  group_by(strain, antibiotic, timepoint) |>
  summarise(
    n = n(),
    pcn_ter_mean = mean(pcn_ter),
    pcn_ter_sd = sd(pcn_ter),
    pcn_ori_mean = mean(pcn_ori),
    pcn_ori_sd = sd(pcn_ori),
    ori_ter_mean = mean(ot),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2)
```
